---
title: "Crime rate falls off before midnight on most days of the week, but stays high later into the night on Fridays and Saturdays"
author: "Scott Piraino"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data and Methods Summary

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE}
require(RCurl)
seattle_url = getURL("https://raw.githubusercontent.com/uwescience/datasci_course_materials/master/assignment6/seattle_incidents_summer_2014.csv")
san_fran_url = getURL("https://raw.githubusercontent.com/uwescience/datasci_course_materials/master/assignment6/sanfrancisco_incidents_summer_2014.csv")
seattle_data = read.csv(text = seattle_url)
san_fran_data = read.csv(text = san_fran_url)
seattle_data$Year_date = substring(as.character(seattle_data$Date.Reported), 7, 10)
seattle_data$Month_date = substring(as.character(seattle_data$Date.Reported), 1, 2)
seattle_data$Day_date = substring(as.character(seattle_data$Date.Reported), 4, 5)
san_fran_data$Year_date = substring(as.character(san_fran_data$Date), 7, 10)
san_fran_data$Month_date = substring(as.character(san_fran_data$Date), 1, 2)
san_fran_data$Day_date = substring(as.character(san_fran_data$Date), 4, 5)
seattle_data$date_numeric = substring(as.character(seattle_data$Date), 1, 10)
seattle_data$Hour_date = as.numeric(substring(as.character(seattle_data$Date), 12, 13))
seattle_data$AM_date = substring(as.character(seattle_data$Date), 21, 22)
seattle_data$Hour_date[seattle_data$AM_date == "PM" & seattle_data$Hour_date != 12] = seattle_data$Hour_date[seattle_data$AM_date == "PM" & seattle_data$Hour_date != 12] + 12
seattle_data$Hour_date[seattle_data$AM_date == "AM" & seattle_data$Hour_date == 12] = 0
san_fran_data$date_numeric = substring(as.character(san_fran_data$Date), 1, 10)
san_fran_data$Hour_date = as.numeric(substring(as.character(san_fran_data$Time), 1, 2))
seattle_limited_time = seattle_data[seattle_data$Year_date == "2014" & seattle_data$Month_date %in% c("06", "07", "08"), ]
san_fran_limited_time = san_fran_data[san_fran_data$Year_date == "2014" & san_fran_data$Month_date %in% c("06", "07", "08"), ]
seattle_limited_time_narc = seattle_limited_time #[seattle_limited_time$Summarized.Offense.Description == "NARCOTICS", ]
san_fran_limited_time_narc = san_fran_limited_time #[san_fran_limited_time$Category == "DRUG/NARCOTIC", ]
san_fran_narc_hour_tab = table(san_fran_limited_time_narc$Hour_date)
seattle_narc_hour_tab = table(seattle_limited_time_narc$Hour_date)
```
This is my attempt to visualize data from criminal incidents in Seattle and San Fransisco for the Coursera Data Science at Scale course. I was interested in how different aspects of time such as time of day or day of the week are related to crime, so I have produced several visualizations that are meant to shed light of these relationships. I'll give a brief explanation of what data I used and how I created the visualizations here, and then I'll jump into the analysis. I retrieved the drime data for Seattle and San Fransisco from the course GitHub page, which includes the data for 2014 for both cities. I only used incidents that are listed as having occured in the months of June, July and August in 2014. To ensure reproducibility, I created this HTML document using R Markdown. The code used to create this analysis is available on GitHub [here](https://github.com/ScottWPiraino/coursera_dsas_projects/blob/master/crime_project/crime_project.Rmd). In order to reproduce the analysis, ensure that all the required R packages (listed at the end of this section) as well as a working version of R Studio are installed, and then use knitr to generate the HTML document from the R Markdown file. This can be easily done in R Studio by loading the .Rmd file into R Studio and clicking the "Knit HTML" button. I have already done this to produce the HTML document that you are currently viewing, in order to avoid the need for external software. I used http://htmlpreview.github.com/ to make the HTML file on GitHub easily viewable. I produced visualizations with the R package ggplot2, and tweeked the parameters to make the graphs a bit more pretty. I also think one further clarification is helpful. At times in this document I'll say things like "crime incidence is lower in the early morning compared to the evening". In all cases, these statements are only intended to apply to this specific dataset. I'm not intending to make any claims about the statistical validity or generalizability of these statements, and they aren't meant to be causal claims. These statements are meant to be descriptive/exploratory only, and are not meant to have any stronger meaning (i.e. statistical significance, predictive, causal).  
  
Required packages: R Studio and the packages that come with it, RCurl, ggplot2, reshape2

## Visualizations and Analysis

```{r hour_plot, echo = FALSE, message = FALSE, warning = FALSE}
require(ggplot2)
require(reshape2)
combined_narc_freq_hour = as.data.frame(cbind(names(san_fran_narc_hour_tab), as.numeric(san_fran_narc_hour_tab)/sum(san_fran_narc_hour_tab)))
names(combined_narc_freq_hour) = c("Hour", "San_Fran")
combined_narc_freq_hour$Seattle = rep.int(0, length(combined_narc_freq_hour$San_Fran))
combined_narc_freq_hour$Seattle[combined_narc_freq_hour$Hour %in% names(seattle_narc_hour_tab)] = as.numeric(seattle_narc_hour_tab)/sum(seattle_narc_hour_tab)
combined_narc_freq_hour[ , 1] = as.numeric(as.character(combined_narc_freq_hour[ , 1]))
combined_narc_freq_hour[ , 2] = as.numeric(as.character(combined_narc_freq_hour[ , 2]))
combined_narc_freq_hour[ , 3] = as.numeric(as.character(combined_narc_freq_hour[ , 3]))
seattle_temp = combined_narc_freq_hour[ , c(1,3)]
names(seattle_temp) = c("Hour", "San_Fran")
combined_narc_freq_hour_long = as.data.frame(rbind(combined_narc_freq_hour[ , c(1,2)], seattle_temp))
combined_narc_freq_hour_long[ , 1] = as.numeric(as.character(combined_narc_freq_hour_long[ , 1]))
combined_narc_freq_hour_long[ , 2] = as.numeric(as.character(combined_narc_freq_hour_long[ , 2]))
names(combined_narc_freq_hour_long) = c("Hour", "Freq")
combined_narc_freq_hour_long$City = c(rep("San Fransisco", 24), rep("Seattle", 24))
print(ggplot(combined_narc_freq_hour_long, aes(x = Hour, y = Freq, group = City, colour = City)) + geom_line() + geom_point() + theme(axis.line.x = element_line(colour = "black"), axis.line.y = element_line(colour = "black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_blank(), panel.background = element_blank(), legend.position = "bottom", axis.ticks = element_line(colour = "black"), axis.text = element_text(colour = "black")) + labs(x = "Hour of day", y = "Fraction of incidents", title = "Figure 1"))
```

For this assignment I chose to explore how the incidence of crime varies with different aspects of time. As a metric for crime incidence I simply chose to use the number of incidents reported in the dataset for all categories of crime. The first thing that I wanted to look into was how crime incidence varies with time of day in both Seattle and San Fransisco. Both datasets come with an indication of when the crime was reported, but since I wanted to visualize these times in a way that allows a comparsion between the two cities I decided to transform them in a way that makes the times comparable. First, I isolated the times and transformed them into 24 hour time so that 12 midnight is represented by a time of 0 and 11 at night is represented by 23. I grouped the times into bins at each hour, so that the first bin from 12 midnight to 1 in the morning has time 0, 1AM to 2AM has time 1, and so on for the times from 0 to 23. Binning in this way allowed me to count the number of times that occured in each bin for each city, expressed as a fraction of total incidents in the city. To visualize the relationship between time of day and crime I then plotted the fraction of incidents against the time of day seperately for both cities is Figure 1. The blue line is Seattle and the red line in San Fransisco. The vertical axis plots the fraction of crimes that occurred at a specific time and the horizontal axis plots time of day (in bin format). Since the time is given in 24-hour time the far left is midnight. Proceeding to the right, we pass the early morning hours, reach noon in the middle, and then go on into the evening, with 11PM (23 on a plot) at the far right. For both cities, crime incidence is moderate around midnight, decreases in the early morning and reaches a low at around 5AM, then increases during the afternoon until it levels off and starts to fall around the middle of the day. Both cities have a very similar overall trend, but Seattle looks like it might peak and begin falling off a little earlier in the day, whereas San Fransisco doesn't see the incidence start to fall until a bit later, although it's not clear whether this is due to noise or not.  
  
```{r day_plot, echo = FALSE, message = FALSE, warning = FALSE}
san_fran_day_of_week = factor(san_fran_limited_time_narc$DayOfWeek)
san_fran_day_of_week_reorder = factor(san_fran_day_of_week, levels(san_fran_day_of_week)[c(4, 2, 6, 7, 5, 1, 3)])
san_fran_day_of_week_tab = table(san_fran_day_of_week_reorder)
san_fran_day_of_week_frame = as.data.frame(cbind(names(san_fran_day_of_week_tab), as.numeric(san_fran_day_of_week_tab)))
names(san_fran_day_of_week_frame) = c("Day", "Count")
san_fran_day_of_week_frame$Count = as.numeric(as.character(san_fran_day_of_week_frame$Count))
san_fran_day_of_week_frame$Frac = san_fran_day_of_week_frame$Count/sum(san_fran_day_of_week_frame$Count)
san_fran_day_of_week_frame$Day = factor(san_fran_day_of_week_frame$Day, levels(san_fran_day_of_week_frame$Day)[c(4, 2, 6, 7, 5, 1, 3)])
print(ggplot(san_fran_day_of_week_frame, aes(x = Day, y = Frac, colour = "red", group = 1)) + geom_line() + geom_point() + theme(axis.line.x = element_line(colour = "black"), axis.line.y = element_line(colour = "black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_blank(), panel.background = element_blank(), legend.position = "none", axis.ticks = element_line(colour = "black"), axis.text = element_text(colour = "black")) + labs(x = "Day of week", y = "Fraction of incidents", title = "Figure 2"))
```

Next, I wanted to look at how crime incidence varies over days of the week. The data for San Fransisco already has the incidents annotated with day of the week, so I performed a similar analysis as above, counting up the number of incidents that occurred at each day of the week and plotting the proportion of incidents on the vertical axis and day of the week on the horizontal, this time using only the San fransisco data. This plot is shown in Figure 2. Friday and Saturday both have higher proportions of incidents compared to the rest of the days, and Sunday has more incidents than Monday - Thursday, but not as much as Friday and Saturday. Wednesday also has a sightly higher number of incidents compared to the other week days.  
  
```{r hour_day_plot, echo = FALSE, message = FALSE, warning = FALSE}
san_fran_day_hour_tab = table(san_fran_limited_time_narc$DayOfWeek, san_fran_limited_time_narc$Hour_date)
san_fran_day_hour_tab_long = as.data.frame(san_fran_day_hour_tab)
names(san_fran_day_hour_tab_long)[1:2] = c("Day", "Hour")
san_fran_day_hour_tab_long$Day = factor(san_fran_day_hour_tab_long$Day, levels(san_fran_day_hour_tab_long$Day)[c(c(4, 2, 6, 7, 5, 1, 3))])
san_fran_day_hour_tab_long$Weekend = rep("Week-day", length(san_fran_day_hour_tab_long$Day))
san_fran_day_hour_tab_long$Weekend[san_fran_day_hour_tab_long$Day %in% c("Friday", "Saturday")] = "Week-end"
print(ggplot(san_fran_day_hour_tab_long, aes(x = Hour, y = Freq, group = Day, linetype = Day, colour = Weekend)) + geom_line() + geom_point() + theme(axis.line.x = element_line(colour = "black"), axis.line.y = element_line(colour = "black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_blank(), panel.background = element_blank(), legend.position = "bottom", axis.ticks = element_line(colour = "black"), axis.text = element_text(colour = "black")) + labs(x = "Hour of day", y = "Number of incidents", title = "Figure 3") + scale_colour_manual(values = c("orange", "purple")))
```  
  
Finally, I wanted to put the two together. Do certain days of the week show a different pattern on the hour to hour level? To assess this possbility, I again used San Fransisco data and counted the number of incidents at each day-hour combination (e.g. hour 0 on Friday, hour 2 on Monday, etc.). In Figure 3 I show essentially the same plot as Figure 1, but this time split up by day of the week. In Figure 3 I plotted absolute number of incidents rather than a proportion because it seems that absolute differences are potentially more meaningful when comparing data from the same city, since factors like total population or differences in enforcement are less worrisome. Each day is plotted using separate line type, and I have also colored the figure so that Friday and Saturday are colored purple, and all other days are colored orange. While most days see the crime incidence fall off on the far right edge of the plot, Friday and Saturday both dip slightly, but then tick back up again so that for these two days specifically crime incidence stays high approaching midnight, in contrast to the other days of the week.